Bayesian Deep Gaussian Processes for Correlated Functional Data: \\ A Case Study in Cosmological Power Spectra

### 2025-04-25 paper/main.tex
update plots, include suggested edits

Computer simulation experiments are invaluable tools in the study of cosmology.
Experiments which simulate the expansion of the universe are growing
in prevalance and complexity \citep[e.g.,][]{lawrence2010coyote,moran2023mira}
\textcolor{red}{(more cites here? possibly those commented out above?)}.  
We are particularly motivated by a simulation
of the matter power spectrum, which describes the distribution of matter as a 
function of spatial scale. 
%It is often represented as a function of wavenumber $k$ (units Mpc$^{-1}$), which is 
%inversely related to spatial scale. 
On large scales, power spectrum behave according to linear perturbation 
theory \citep{pietroni2008flowing, lesgourgues2009non}.  On smaller scales, nonlinear 
dynamics require the use of computationally intensive simulations, yielding
correlated functional data.
%the Coyote Universe \citep{lawrence2010coyote} and the Mira-Titan Universe \citep{moran2023mira} 
%are two simulation suites dedicated to this effort.  
Our overarching goal is to leverage such simulation data to 
first estimate and then predict underlying power matter spectra for various 
cosmological parameterizations.

@@

Our second objective is to leverage simulation data from a limited set of
cosmologies in order to predict power matter spectra for unobserved cosmologies.
Due to the computationally expensive nature of the Mira-Titan simulation suite
\textcolor{red}{(do we know how long it takes to run?)},
it is infeasible to evaluate the simulation for every possible eight-dimensional
cosmological configuration which may be of interest.  Instead, we desire
a statistical ``emulator'' or ``surrogate model'' 
\citep{santner2003design,gramacy2020surrogates} which will provide quick and effective
predictions of power matter spectra for any cosmological parameterization.
\citet{moran2023mira} provide a state-of-the-art emulator for the Mira-Titan
simulation suite, which was trained on 117 simulations, and is termed 
the ``Cosmic Emu.''\footnote{\url{https://github.com/lanl/CosmicEmu}}
We leverage the same training data, in conjunction with our Bayesian hierarchical
DGP model and basis function representations, to train a Gaussian process surrogate 
on principal component weights in order to predict spectra for unobserved
cosmologies \citep{higdon2008computer, higdon2010estcosmo}. 
Our method compares favorably to Cosmic Emu on held-out cosmologies.

@@

\begin{figure}[ht]
    \centering
    \includegraphics[width=\linewidth, trim=0 13 0 45, clip=TRUE]{plot_data.png}
    \caption{{\it Left:} The perturbation theory (solid green), low resolution runs (solid grey), 
    and high resolution run  (solid pink) for a particular cosmology. 
    The weighted average of these is shown in dashed orange. Dotted lines at the bottom 
    indicate where each data type is deemed approximately unbiased. 
    {\it Right:} Same, but restricted to wavenumber ($k$) values where the low resolution 
    runs are approximately unbiased.
    \textcolor{red}{(Either make one legend for both graphs or ditch the legend 
    altogether and narrate the colors/line types in the caption.  If legend stays, "ind" needs
    to go.  Pink line is too hard to see.  Make gray lines thinner and pink thicker?
    Since later we will use a centered version, we could add a 3rd panel here that shows the
    same data centered, so the reader can understand the connection to the plots later.)}}
    \label{fig:plot_data}
\end{figure}

These spectra are represented as a function of wavenumber $k$.  Following 
\cite{moran2023mira}, we model on the emulation space 
$\mathcal{P}(k)=\log_{10}\left(\frac{k^{1.5}P(k)}{2\pi^2}\right)$, where $P(k)$ represents the
original spectra. Output is available for $n=351$ values of $k$, spanning
$0.001 \leq k \leq 5$, but each data type has unique wavenumber ranges
where it is deemed approximately unbiased. For $k<0.04$, only $y_p$ provides a reliable 
estimate of the true spectra. $y_{\ell_r}$ are valid for $0.04 \leq k \leq 0.25$, and 
$y_h$ is valid for $0.04 \leq k \leq 5$.  Figure \ref{fig:plot_data} shows an example 
of the output for a particular cosmology. \textcolor{red}{(Add some narration here?  Something
that doesn't repeat the caption but complements it.)}

From the multiple low- and high-resolution spectra across all cosmologies, 
\cite{moran2023mira} obtained precision estimates of these spectra across the $n$ 
wavenumber values ($p_1,\dots,p_n$) using a log-log regression model, and a multiplier 
$c$ for the increase in precision from the low- to high- resolution output. 
\textcolor{red}{(I'm missing a bit of context here - why did we need the precision
estimates in the first place?  Can we show a graph of $p_1,\dots, p_n$ to illustrate
how these change over $k$ values?)}
Given that the perturbation theory spectra is non-stochastic, we treat this as having a 
precision of $10^5$. We leverage these precision estimates within our model at each 
data type's unbiased wavenumber values to define 
$\Lambda_p = \text{diag}\left((10^5)_{k < 0.04}\right)$, 
$\Lambda_{\ell} = \text{diag}\left(16(p_1,\dots,p_n)_{0.04 \leq k < 0.25}\right)$, 
$\Lambda_h = \text{diag}\left(c(p_1,\dots,p_n)_{0.04 \leq k \leq 5}\right)$. 
\textcolor{red}{(Unless we are really short on space, could we show these in
actual "bmatrix" form?  Or visually somehow?)}
Here, each precision matrix is diagonal and the diagonal elements consist of the 
precision values or zero, depending on whether the $k$ value is approximately unbiased 
for that data type. With this, we calculate a weighted average for each cosmology: 
$\bar y = \Lambda^{-1}(\Lambda_p y_p + \Lambda_h y_h + \Lambda_{\ell} \bar{y}_\ell)$, 
where $\Lambda = \Lambda_p + \Lambda_\ell + \Lambda_h$ is the precision matrix for 
the weighted average for a particular cosmology. \textcolor{red}{(Revisit the figure and
comment on the weighted average.  Explain why the weighted
average on its own is not an effective estimate.  Motivate why we need a more complex model.)}

@@

\begin{figure}
    \centering
    \includegraphics[width=6in]{CAMB_fit_model1.jpg}
    \caption{Example of model fit for the first model from the CAMB data, with LOESS
             mean term subtracted. \textcolor{red}{Currently, we can compare the 
             orange lines (posterior mean and 95\% credible interval) to the blue line,
             representing the infinite-resolution ``true" spectrum. Steve: Update this plot to be
             more clean, use different line types: dotted, dash, solid, etc.}}   
    \label{fig:fit_camb}
\end{figure}

We use 32 CAMB batches to train the DGP.HM model and reserve the other 32 as a test set.
To imitate the role of the perturbation theory output of Mira-Titan, we anchor estimates of the CAMB
power spectrum at the infinite-resolution for $\log_{10}(k) < -1.7$, and then rely 
strictly on the low- and high-resolution runs for $k$ above this. We take a weighted
average of the CAMB runs and assume heteroskedastic errors in a similar fashion, now with
no spatial dependence (compare low-resolution runs of Figure \ref{fig:fit_camb} 
with Figure \ref{fig:plot_data}).

>>>>

Computer simulation experiments are invaluable tools in the study of cosmology.
Experiments which simulate the expansion of the universe are growing
in prevalance and complexity \citep[e.g.,][]{lawrence2010coyote,derose2019aemulus,
nishimichi2019dark,angulo2021bacco,euclid2021euclid,moran2023mira}.  
We are particularly motivated by a simulation
of the matter power spectrum, which describes the distribution of matter as a 
function of spatial scale. 
%It is often represented as a function of wavenumber $k$ (units Mpc$^{-1}$), which is 
%inversely related to spatial scale. 
On large scales, power spectrum behave according to linear perturbation 
theory \citep{pietroni2008flowing, lesgourgues2009non}.  On smaller scales, nonlinear 
dynamics require the use of computationally intensive simulations, yielding
correlated functional data.
%the Coyote Universe \citep{lawrence2010coyote} and the Mira-Titan Universe \citep{moran2023mira} 
%are two simulation suites dedicated to this effort.  
Our overarching goal is to leverage such simulation data to 
first estimate and then predict underlying power matter spectra for various 
cosmological parameterizations.

@@

Our second objective is to leverage simulation data from a limited set of
cosmologies in order to predict power matter spectra for unobserved cosmologies.
Due to the computationally expensive nature of the Mira-Titan simulation suite
(one batch of simulations can take multiple weeks to run on a supercomputer),
it is infeasible to evaluate the simulation for every possible eight-dimensional
cosmological configuration which may be of interest.  Instead, we desire
a statistical ``emulator'' or ``surrogate model'' 
\citep{santner2003design,gramacy2020surrogates} which will provide quick and effective
predictions of power matter spectra for any cosmological parameterization.
\citet{moran2023mira} provide a state-of-the-art emulator for the Mira-Titan
simulation suite, which was trained on 117 simulations, and is termed 
the ``Cosmic Emu.''\footnote{\url{https://github.com/lanl/CosmicEmu}}
We leverage the same training data, in conjunction with our Bayesian hierarchical
DGP model and basis function representations, to train a Gaussian process surrogate 
on principal component weights in order to predict spectra for unobserved
cosmologies \citep{higdon2008computer, higdon2010estcosmo}. 
Our method compares favorably to Cosmic Emu on held-out cosmologies.

@@

\begin{figure}[ht]
    \centering
    \includegraphics[width=\linewidth, trim=0 13 0 45, clip=TRUE]{plot_data.png}
    \caption{{\it Left:} The perturbation theory (solid green), low resolution runs (solid grey), 
    and high resolution run  (solid pink) for a particular cosmology. 
    The weighted average of these is shown in dashed orange. Dotted lines at the bottom 
    indicate where each data type is deemed approximately unbiased. 
    {\it Middle:} Same as left, but restricted to wavenumber ($k$) values where the low resolution 
    runs are approximately unbiased.
    {\it Right:} Same as left, but with estimated posterior mean subtracted.
    \textcolor{red}{Rerun on Windows for $\mathcal{P}$}}
    \label{fig:plot_data}
\end{figure}

These spectra are represented as a function of wavenumber $k$.  Following 
\cite{moran2023mira}, we model on the emulation space 
$\mathcal{P}(k)=\log_{10}\left(\frac{k^{1.5}P(k)}{2\pi^2}\right)$, where $P(k)$ represents the
original spectra. Output is available for $n=351$ values of $k$, spanning
$0.001 \leq k \leq 5$, but each data type has unique wavenumber ranges
where it is deemed approximately unbiased. For $k<0.04$, only $y_p$ provides a reliable 
estimate of the true spectra. $y_{\ell_r}$ are valid for $0.04 \leq k \leq 0.25$, and 
$y_h$ is valid for $0.04 \leq k \leq 5$.  Figure \ref{fig:plot_data} shows an example 
of the output for a particular cosmology. When we focus on $k$ values where $y_{\ell_r}$ 
is valid, we see each low-resolution and the high-resolution run oscillate around 
a mean term, which is relatively smooth, compared to these realizations. When such a 
mean term is subtracted, we will obtain a plot like the right panel in 
Figure \ref{fig:plot_data}.

From the multiple low- and high-resolution spectra across all cosmologies, 
\cite{moran2023mira} obtained precision estimates of these spectra across the $n$ 
wavenumber values ($p_1,\dots,p_n$) using a log-log regression model, and a multiplier 
$c$ for the increase in precision from the low- to high- resolution output. 
We can easily incorporate these estimates in our model to account for nonstationarity 
(i.e., non-constant variance) of the runs. We see as $k$ increases, the runs' precision 
increases in the middle panel of Figure \ref{fig:plot_data}). 
Given that the perturbation theory spectra is non-stochastic, we treat this as having a 
precision of $10^5$. We leverage these precision estimates within our model at each 
data type's unbiased wavenumber values to define the following diagonal precision matrix:
% $\Lambda_p = \text{diag}\left((10^5)_{k < 0.04}\right)$, 
% $\Lambda_{\ell} = \text{diag}\left(16\times(p_1,\dots,p_n)_{0.04 \leq k < 0.25}\right)$, 
% $\Lambda_h = \text{diag}\left(c\times(p_1,\dots,p_n)_{0.04 \leq k \leq 5}\right)$.

\[
\Lambda = 
\begin{bmatrix}
\Lambda_p & 0 & 0 \\
0 & \Lambda_\ell & 0 \\
0 & 0 & \Lambda_h
\end{bmatrix}
=
\begin{bmatrix}
\operatorname{diag}(10^5, \ldots, 10^5) &  &  \\
 & \operatorname{diag}(16\times(p_1, \ldots, p_n)) &  \\
 &  & \operatorname{diag}(c\times(p_1, \ldots, p_n))
\end{bmatrix}
\]

\[
\text{where } 
\begin{cases}
\Lambda_p \text{ for } k < 0.04 \\
\Lambda_\ell \text{ for } 0.04 \leq k < 0.25 \\
\Lambda_h \text{ for } 0.25 \leq k \leq 5
\end{cases}
\]

Above, each matrix element is either the corresponding precision value or zero, 
depending on whether the $k$ value is approximately unbiased for that data type. 
With this, we calculate a weighted average for each cosmology: 
$\bar y = \Lambda^{-1}(\Lambda_p y_p + \Lambda_h y_h + \Lambda_{\ell} \bar{y}_\ell)$, 
where $\Lambda = \Lambda_p + \Lambda_\ell + \Lambda_h$ is the precision matrix for 
the weighted average for a particular cosmology. On its own, the weighted average
has too much variation baked in from the limited number of model runs. There is also
no spatial dependence accounting for the correlated errors seen in Figure \ref{fig:plot_data}.
To account for these modeling challenges, we need a more complex model.

@@

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{CAMB_fit_model1.jpg}
    \caption{Example of DGP.HM model fit for the first batch of model runs from the CAMB data, 
             with LOESS mean term subtracted. We can compare the orange lines 
             (posterior mean and dotted 95\% credible interval) to the light blue line,
             representing the infinite-resolution ``true" spectrum.}   
    \label{fig:fit_camb}
\end{figure}

We use 32 CAMB batches to train the DGP.HM model and reserve the other 32 as a test set.
To imitate the role of the perturbation theory output of Mira-Titan, we anchor estimates of the CAMB
power spectrum at the infinite-resolution for $\log_{10}(k) < -1.7$, and then rely 
strictly on the low- and high-resolution runs for $k$ above this. We take a weighted
average of the CAMB runs and assume heteroskedastic errors in a similar fashion. For the CAMB
data, lengthscale estimates are consistently around $3\times 10^{-7}$, evidence that we
can simplify our covariance structure to contain no spatial dependence 
(compare low-resolution runs of Figure \ref{fig:fit_camb} 
with Figure \ref{fig:plot_data}).
